Parsing Postfix log files

### 2008-08-25 doc/logparser-implementation.tex
Changes made following suggestions from diction.

    \item Postfix logs sufficient information to make it possible to
        accurately reconstruct the actions it has taken.  There are a
        number of heuristics used when parsing; see
        \sectionref{identifying-bounce-notifications},
        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
        logging after cleanup} for details.

@@

In some ways this task is similar to reverse engineering or replicating a
black box program based solely on its inputs and outputs.  Although the
source code is available,\footnote{Reading and understanding the source
code would require a significant investment of time: there are 375,750
lines of code, documentation, etc.\ in Postfix 2.5.1's 17MB of source
code.} there are advantages to treating Postfix as a black box while
developing the parser:

@@

    \item The parser acts as a second source of information, with the
        information gathered from empirical evidence.  An interesting
        project would be to compare the empirical knowledge inherent in the
        parsing algorithm with the documentation and source code of
        Postfix.

@@

The database schema can be conceptually divided in two: the rules, which
are used to parse log files, and the data saved from the parsing of log
files.  Rules have the fields required to parse the log lines, extract data
to be saved, and the action to be executed; they also have several fields
which aid the user in understanding the meaning of the log lines parsed by
each rule.  The rules are described in detail in \sectionref{rules} but the
fields are covered in \sectionref{rule attributes}.

@@

An important but easily overlooked benefit of storing the rules in the
database is the connection between rules and results: if more information
is required when examining a result, the rule which produced the result is
available for inspection --- each result references the rule which created
it.  There is no ambiguity about which rule resulted in a particular
result, eliminating one potential source of confusion.

@@

    \item [id] A unique identifier for each rule which other tables can use
        to refer to a specific rule.

@@

    \item [restriction\_name] The restriction which caused the mail to be
        rejected.  Only applicable to rules which have an action of
        \texttt{REJECTION}, other rules should have an empty string.

@@

            \item [BOUNCED] The mail has bounced, due to a mail loop,
                delivery failure, or five day timeout.

@@

            \item [INFO] Represents an unspecified intermediate action that
                the parser is not interested in per se, but which does log
                useful information, supplementing other log lines.

            \item [IGNORED] An action which is not only uninteresting in
                itself, but which also provides no useful data.

            \item [POSTFIX\_RELOAD] The administrator has instructed
                Postfix to start or stop, and all existing \daemon{smtpd}
                processes will be terminated.  This does not negatively
                impact on the log files or mail queued by Postfix for
                delivery.

@@

    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
        log lines the rule applies to.  This avoids needlessly trying rules
        which will not match the log line, or worse, might match
        unintentionally.  Rules whose program is \texttt{*} will be tried
        against any log lines which are not parsed by program specific
        rules.

@@

        For efficiency the keywords are expanded and every rule's \regex{}
        is compiled before attempting to parse the log file --- otherwise
        each \regex{} would be recompiled each time it was used, resulting
        in a large, data dependent slowdown.  Efficiency concerns are
        discussed in \sectionref{parser efficiency}, with the impact of
        caching compiled \regexes{} covered in \sectionref{Caching compiled
        regexes}.

    \item [result\_cols, connection\_cols] Specifies how the fields in the
        log line will be extracted.  The format is:
        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
        \newline i.e.\ semi-colon or comma separated assignment statements,
        with the variable name on the left and the matching capture from
        the \regex{} on the right hand side.  Data specified by
        \texttt{result\_cols} will be entered in the \texttt{results}
        table, while \texttt{connection\_cols} populates the
        \texttt{connection} table.  The list of acceptable variable names
        is:

@@

    \item [result\_data, connection\_data] Sometimes rules need to supply a
        piece of data which is not present in the log line: e.g.\ setting
        \texttt{smtp\_code} when mail is accepted.  The format and allowed
        variables are the same as for \texttt{result\_cols} and
        \texttt{connection\_cols}, except that arbitrary
        data\footnote{Commas and semi-colons cannot be escaped and thus
        cannot be used.  This feature is intended for use with small
        amounts of data rather than large amounts in any one rule, so
        dealing with escape sequences was deemed unnecessary.} is permitted
        on the right hand side of the assignment.

@@

    \item [queueid] Specifies the capturing field in the \regex{}
        containing the queueid, or zero if the log line does not contain a
        queueid.  Many log lines will not contain a queueid, e.g.\
        rejections logged before a mail has been accepted (a queueid will
        not have been allocated), or log lines which are not tied to one
        particular mail.

@@

    \item [helo] The hostname used in the HELO command.  The HELO hostname
        occasionally changes during a connection, presumably because spam
        or virus senders think it is a good idea.  By default Postfix only
        logs the HELO hostname when it rejects an \SMTP{} command, but it
        is quite easy to rectify this, as described in \sectionref{logging
        helo}.

@@

Every interesting\footnote{All log lines are interesting unless the
postfix\_action of the rule which parses them is \texttt{INFO} or
\texttt{IGNORE}, as described in \sectionref{actions-in-detail}.} log line
will have an entry in the results table, e.g.\ rejecting an \SMTP{}
command, delivering a mail, or bouncing a mail (see
\sectionref{postfix_action} for typical values of
\texttt{postfix\_action}).  Each row is associated with a single
connection, though there may be many results per connection.

@@

    \item [rule\_id] The id of the entry in the rules table which matched
        the log line and created this result.

@@

The table containing the rules used by the parser and both tables
containing the data extracted from the Postfix log files were described,
with the purpose of each field discussed in detail.  A clear,
comprehensible schema is essential when using the extracted data; it is
more important when using the data than when storing it, because storing
the data is a write-once operation, whereas utilising the data requires
frequent searching, sorting and manipulation of the data to produce
customised reports and/or statistics.

@@

The example rule below would match the following log line:

@@

    \item [program and regex:] If the program in the rule equals the
        program which logged the log line, a match using the rule's
        \regex{} will be attempted against the log line; if the match is
        successful the action will be executed, if not the next rule will
        be tried.  If the program-specific rules do not match the log line,
        the parser will fall back to generic rules; if those rules are
        unsuccessful a warning will be issued.

@@

    \item [queueid:] The index of the capture in the \regex{} which
        supplies the queueid, or zero if the log line does not contain a
        queueid.  This allows the correct mail can be found by queueid and
        actions performed on it.

@@

The log files produced by Postfix differ from installation to installation,
because administrators have the freedom to choose the subset of available
restrictions which suits their needs, including using different \DNSBL{}
services, policy servers, or custom rejection messages.  To facilitate easy
parsing of new log lines, the parser's design separates parsing rules from
parsing actions: adding new actions is difficult, but adding new rules to
parse new rejection messages is trivial, and also occurs much more
frequently.  The implementation provides a program to ease the process of
creating new rules from unparsed log lines, based on the algorithm
developed by Risto Vaarandi~\cite{risto-vaarandi} for his
\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
be outlined as part of the general explanation below.

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained while writing rules and performs a large number of
        substitutions on the log input lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \IP{} addresses, the
        standard start of rejection messages, etc.) with \regex{} keywords
        which the parser will expand when it loads the rule (see the
        \regex{} entry in \sectionref{rule attributes}).  The purpose of
        this phase is to utilise existing knowledge to create more accurate
        \regexes{}.  The new log lines are written to a temporary file,
        which all subsequent stages use instead of the original input file.

@@

        The original algorithm only maintains counters for words whose hash
        result from the previous phase has a high frequency; this reduces
        the number of counters maintained by the algorithm, reducing the
        memory requirements of the algorithm.

        As time goes on the amount of memory typically available increases
        and the requirement to reduce memory requirements decreases, hence
        the modified algorithm omits the hashing step and maintains
        counters for all tokens. In addition the majority of unique or
        infrequently occurring words will have been substituted with
        keywords during the first phase, vastly reducing the number of
        tuples to maintain counters for.

@@

    \item [Build regexes]  The words are reassembled to produce a \regex{}
        matching the log line, and a counter maintained for each \regex{}.
        Contiguous sequences of \texttt{.+} in the newly reassembled
        \regexes{} are collapsed to a single \texttt{.+}; any resulting
        duplicate \regexes{} are combined, and their counters added
        together.  If the \regex{}'s counter is lower than the threshold
        supplied by the user the \regex{} is discarded; this second
        threshold is independent of the threshold used to differentiate
        between fixed and variable words, but once again 1\% of input log
        lines is a good starting point; obviously the threshold depends on
        the number and type of input log lines.  The new \regexes{} are
        displayed for the user to add to the database, either as new rules
        or merged into the \regexes{} of existing rules; the counter for
        each \regex{} is also displayed, giving the user an indication of
        how many of the input log lines that \regex{} should match.
        Discarding \regexes{} will result in some of the input log lines
        not being matched; the utility should be run again once the
        unmatched log lines have been isolated by parsing the original log
        files once the parser has been augmented with the new rules.

@@

A second utility is also provided which reads a list of new \regexes{} and
the input given to the first utility.  It tries to match each input log
line against each \regex{}, counting the number of log lines which match
each \regex{}, warning the user if an input log line is matched by more
than one \regex{}, and additionally warning if an input log line is not
matched by any \regex{}.  It displays a summary of how many input log lines
each \regex{} matched, comparing it to the expected number of matches; this
provides the user with an easy method of checking if the \regexes{}
produced by the first utility are correctly matching the input log lines
they are based upon.  A future version of this utility will also group
input log lines by \regex{}, so the user can tweak the \regexes{} if
required.

@@

Whereas the rules are quite simple and each rule is completely independent
of the other rules, the algorithm is significantly more complicated and
highly internally interdependent.  The algorithm deals with all the
complications of parsing, the eccentricities and oddities of Postfix log
files, presenting the resulting data in a normalised, easy to use
representation.  The algorithm's task is to follow the journey each mail
takes through Postfix, combining the data extracted by rules into a
coherent whole, saving it in a useful and consistent form, and performing
housekeeping duties.

@@

    \item The first set of complications encountered: initially obvious
        difficulties which had to be overcome.

@@

    \item Additional complications which have arisen during the development
        of this parser are documented, the solutions to those
        complications, and where those solutions are implemented in the
        algorithm

@@

\Figureref{flow chart image part 1} and \figureref{flow
chart image part 2} show the paths the data representing a mail can take
through the parser algorithm.  The flow chart covers the most common paths
only; there are additional, uncommon paths which are excluded for the sake
of clarity; details of the deviations can be found in
\sectionref{additional complications}.

@@

    \item The last alternative is that the mail is a parent mail (action:
        COMMIT\@; flowchart:~21).  Regardless of the state of its children
        its data is cleaned up and entered in the database (flowchart:~22).
        The parent may have children which are waiting to be entered in the
        database (flowchart:~23); each of those children's data is cleaned
        up and entered in the database, then deleted from the state tables.
        The parent may also have outstanding children which are not yet
        delivered (flowchart:~24), in which case the parent must wait for
        those children to be finished with.  As soon as the last child is
        deleted from the state tables the parent will also be finished with
        (flowchart:~25), and deleted from the state tables.

@@

The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
which is far more complicated than the behaviour explicitly encoded in the
algorithm.  The top-level parser could be written in pseudo-code like so
(with indentation level indicating flow control):

@@

    \item Four implicit branches, where the transition is determined by the
        next log line which is processed for that mail, e.g.\ state 1 can
        transition to either state 4 or state 5.  (1--\{4,5\},
        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})

@@

            \item [host] The hostname of the server which logged the line.

            \item [program] The name of the program which logged the line.

            \item [pid] The \pid{} of the program which logged the line.

@@

    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
        to deal with unusual situations --- after five hours the timer will
        expire and the \daemon{smtpd} will exit.  This occurs very
        infrequently, as there are many other timeouts which should occur
        in the intervening hours: \DNS{} timeouts, timeouts reading data
        from the client, etc\@.  The active connection for that
        \daemon{smtpd} must be cleaned up, entered in the database, and
        deleted from the state tables.

@@

Adding new actions is not as simple as adding new rules, though care has
been taken in the parser's design and implementation to make adding new
actions as painless as possible.  The implementor writes a subroutine which
accepts the standard arguments given to actions, and registers it as an
action\footnote{The new action must be registered before the rules are
loaded, as it is an error for a rule to reference an unregistered action;
this helps catch mistakes made when adding new rules.} by calling the
framework subroutine add\_actions() with the name of the new action
subroutine as a parameter.  No other work is required on the implementor's
part to integrate the action into the parser; all of their attention and
effort can be focused on the correctness of their action.  The only
negative aspect is that the process involves editing the parser source
code, which makes upgrading to a later version of the parser more
difficult, though by no means impossible.  If the author of the new action
wishes, they can take advantage of the parser's object oriented
implementation~\cite{Wikipedia-object-orientation} by subclassing it and
implementing their changes in the derived class, allowing future upgrading
of the parser with greatly reduced chance of conflicts.\footnote{The real
difference between the two approaches is where the new code is placed.  The
simpler option is to change the parser code directly, but those changes
will then have to be made to subsequent versions of the parser, and as the
scope of the changes increases so does the chance of conflict, or mistakes
when copying the action.  The more time consuming option is to write a
subclass containing the new actions and change the program which invokes
the parser so that it uses the subclass rather than the parser; the changes
required to the program invoking the parser are minor and much less likely
to lead to conflicts when upgrading to future versions of the parser.  An
alternative is to submit new actions to the author of the parser for
inclusion in future versions, resulting in two benefits: the new actions do
not need to be maintained separately, and other users of the parser will
avail of the new functionality.} The action may need to extend the list of
valid combinations described in \sectionref{out of order log lines} if the
addition creates a different set of acceptable programs, but this is
unlikely to occur, as it would require parsing log lines from Postfix
components the parser currently ignores.\footnote{The mail server used for
development does not utilise either the \daemon{lmtp} or \daemon{virtual}
delivery agents, so this parser does not have rules to handle log lines
from those components.  Adding new rules to parse those component's log
lines is a simple process, though if their behaviour differs significantly
from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
required.  The mail server in question is a production mail server handling
mail for a university department; the benefit is that the log files used
exhibit the idiosyncrasies and peculiarities a mail server in the wild must
deal with, but the downside is that significantly altering the
configuration just to log messages from a different Postfix component is
not an option.}

@@

This section presented the core of the parser, starting with a very high
level view and the initial complications which arose.  A flow chart showing
the paths a mail may take through the nascent simplified algorithm was
provided, followed by an explanation of those paths, and a discussion of
the parser's emergent behaviour --- the data from the log files creates the
paths in the flow chart, they are not specified anywhere in the parser.
The framework which holds the parser together was covered next, after which
came a description of the current actions provided by the parser, and the
algorithm for analysing unparsed log lines to create \regexes{} for new
rules.  Detecting, diagnosing and defeating complications forms the largest
single portion of this document, mirroring the development of the parser,
The complications are described in the order they were overcome, with
subsequent problems affecting fewer mails (often by an order or magnitude),
though the time required to solve problems increased with each successive
problem.

@@

Postfix 2.2.x (and presumably previous versions) lacks explicit logging
when bounce notifications are generated; suddenly there will be log entries
for a mail which lacks an obvious source.  There are similarities to the
problem of re-injected mails discussed in \sectionref{tracking re-injected
mail}, but unlike the solution described therein bounce notifications do
not eventually have a log line which identifies their source.  Heuristics
must be used to identify bounce notifications, and those heuristics are:

@@

Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
client aborts the first delivery attempt after the first recipient is
accepted, then makes a second delivery attempt for the same recipient which
it continues with until delivery is complete.\footnote{Microsoft Outlook is
one client which behaves in this fashion; no attempt has been made to
identify other clients which act in a similar way.}  An example dialogue
exhibiting this behaviour is presented below (lines starting with a three
digit number are sent by the server, the other lines are sent by the
client):

@@

    \item If one or more recipients have been accepted Postfix will have
        allocated a queueid for the incoming mail, and there will be a mail
        in the state tables which needs to be dealt with.

@@

The author has only observed this complication occurring after a timeout,
though there may be other circumstances which trigger it.  Sometimes the
\daemon{cleanup} line is logged after the timeout line; parsing this line
causes the creation of a new state table entry for the queueid in the log
line.  This is incorrect because the line actually belongs to the mail
which has just been discarded; the next log line for that queueid will be
seen when the queueid is reused for a different mail, causing a queueid
clash and the appropriate warning.

@@

The solution to this problem is to examine the list of programs which have
logged lines for each mail, comparing the list against a table of
known-good combinations of programs (this check is performed during the
COMMIT action).  If the mail's combination is found in the valid list the
mail can be entered in the database; if the combination is not found entry
must be postponed and the mail flagged for later entry.  The
SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it is
found; if the additional log lines have caused the mail to reach a valid
combination entry will proceed, otherwise it must be postponed once more.

@@

    \item It appears that the hostname used in the HELO command is not
        logged if the mail is accepted.\footnote{Tested with Postfix
        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
        2.5. XXX TEST THIS AGAIN.}  Rectifying this is relatively simple:
        create a restriction which is guaranteed to warn for every accepted
        mail, as follows:

>>>>

    \item Postfix logs enough information to make it possible to accurately
        reconstruct the actions it has taken.  There are several heuristics
        used when parsing; see
        \sectionref{identifying-bounce-notifications},
        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
        logging after cleanup} for details.

@@

In some ways this task is similar to reverse engineering or replicating a
black box program based solely on its inputs and outputs.  Although the
source code is available,\footnote{Reading and understanding the source
code would require a significant investment of time: there are 375,750
lines of code, documentation, etc.\ in Postfix 2.5.1's 17MB of source
code.} there are advantages to treating Postfix as a black box during
parser development:

@@

    \item The parser acts as a second source of information, with the
        information gathered from empirical evidence.  A separate project
        could compare the empirical knowledge inherent in the parsing
        algorithm with the documentation and source code of Postfix to see
        how closely the two agree.

@@

The database schema can be conceptually divided in two: the rules, which
are used to parse log files, and the data saved from the parsing of log
files.  Rules have the fields required to parse the log lines, extract data
to be saved, and the action to be executed; they also have several fields
that aid the user in understanding the meaning of the log lines parsed by
each rule.  The rules are described in detail in \sectionref{rules} but the
fields are covered in \sectionref{rule attributes}.

@@

An important but easily overlooked benefit of storing the rules in the
database is the connection between rules and results: if more information
is required when examining a result, the rule that produced the result is
available for inspection --- each result references the rule that created
it.  There is no ambiguity about which rule resulted in a particular
result, eliminating one potential source of confusion.

@@

    \item [id] A unique identifier for each rule that other tables can use
        to refer to a specific rule.

@@

    \item [restriction\_name] The restriction that caused the mail to be
        rejected.  Only applicable to rules that have an action of
        \texttt{REJECTION}, other rules should have an empty string.

@@

            \item [BOUNCED] The mail has bounced, because of a mail loop,
                delivery failure, or five day timeout.

@@

            \item [INFO] Represents an unspecified intermediate action that
                the parser is not interested in per se, but that does log
                useful information, supplementing other log lines.

            \item [IGNORED] An action that is not only uninteresting in
                itself, but that also provides no useful data.

            \item [POSTFIX\_RELOAD] The administrator has instructed
                Postfix to start or stop, and all existing \daemon{smtpd}
                processes will be terminated.  This does not negatively
                affect the log files or mail queued by Postfix for
                delivery.

@@

    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
        log lines the rule applies to.  This avoids needlessly trying rules
        that will not match the log line, or worse, might match
        unintentionally.  Rules whose program is \texttt{*} will be tried
        against any log lines that are not parsed by program specific
        rules.

@@

        For efficiency the keywords are expanded and every rule's \regex{}
        is compiled before attempting to parse the log file --- otherwise
        each \regex{} would be recompiled each time it was used, resulting
        in a large, data dependent slowdown.  Efficiency concerns are
        discussed in \sectionref{parser efficiency}, with caching compiled
        \regexes{} covered in \sectionref{Caching compiled regexes}.

    \item [result\_cols, connection\_cols] Specifies how the fields in the
        log line will be extracted.  The format is: \newline \tab{}
        \texttt{smtp\_code = 1; recipient = 2, sender = 4;} \newline i.e.\
        semi-colon or comma separated assignment statements, with the
        variable name on the left and the matching capture from the
        \regex{} on the right hand side.  Data specified by
        \texttt{result\_cols} will be entered in the \texttt{results}
        table, whereas \texttt{connection\_cols} populates the
        \texttt{connection} table.  The list of acceptable variable names
        is:

@@

    \item [result\_data, connection\_data] Sometimes rules need to supply a
        piece of data that is not present in the log line: e.g.\ setting
        \texttt{smtp\_code} when mail is accepted.  The format and allowed
        variables are the same as for \texttt{result\_cols} and
        \texttt{connection\_cols}, except that arbitrary
        data\footnote{Commas and semi-colons cannot be escaped and thus
        cannot be used.  This feature is intended for use with small
        amounts of data rather than large amounts in any one rule, so
        dealing with escape sequences was deemed unnecessary.} is permitted
        on the right hand side of the assignment.

@@

    \item [queueid] Specifies the capturing field in the \regex{}
        containing the queueid, or zero if the log line does not contain a
        queueid.  Many log lines will not contain a queueid, e.g.\
        rejections logged before a mail has been accepted (a queueid will
        not have been allocated), or log lines that are not tied to one
        particular mail.

@@

    \item [helo] The hostname used in the HELO command.  The HELO hostname
        occasionally changes during a connection, presumably because spam
        or virus senders think it is a good idea.  By default Postfix only
        logs the HELO hostname when it rejects an \SMTP{} command, but it
        is quite simple to rectify this, as described in
        \sectionref{logging helo}.

@@

XXX REWRITE THE SENTENCE AND FOOT NOTE\@.

Every interesting\footnote{All log lines are interesting unless the
postfix\_action of the rule that parses them is \texttt{INFO} or
\texttt{IGNORE}, as described in \sectionref{actions-in-detail}.} log line
will have an entry in the results table, e.g.\ rejecting an \SMTP{}
command, delivering a mail, or bouncing a mail (see
\sectionref{postfix_action} for typical values of
\texttt{postfix\_action}).  Each row is associated with a single
connection, though there may be many results per connection.

@@

    \item [rule\_id] The id of the entry in the rules table that matched
        the log line and created this result.

@@

The table containing the rules used by the parser and both tables
containing the data extracted from the Postfix log files were described,
with the purpose of each field discussed in detail.  A clear,
comprehensible schema is essential when using the extracted data; it is
more important when using the data than when storing it, because storing
the data is a write-once operation, whereas utilising the data requires
frequent searching, sorting and manipulation of the data to produce
customised reports or statistics.

@@

The example rule below matches the following log line:

@@

    \item [program and regex:] If the program in the rule equals the
        program that logged the log line, a match using the rule's
        \regex{} will be attempted against the log line; if the match is
        successful the action will be executed, if not the next rule will
        be tried.  If the program-specific rules do not match the log line,
        the parser will fall back to generic rules; if those rules are
        unsuccessful a warning will be issued.

@@

    \item [queueid:] The index of the capture in the \regex{} that
        supplies the queueid, or zero if the log line does not contain a
        queueid.  This allows the correct mail to be found by queueid and
        actions performed on it.

@@

The log files produced by Postfix differ from installation to installation,
because administrators have the freedom to choose the subset of available
restrictions which suits their needs, including using different \DNSBL{}
services, policy servers, or custom rejection messages.  To facilitate
parsing new log lines, the parser's design separates parsing rules from
parsing actions: adding new actions is difficult, but adding new rules to
parse new rejection messages is trivial, and also occurs much more
frequently.  The implementation provides a program to ease the process of
creating new rules from unparsed log lines, based on the algorithm
developed by Risto Vaarandi~\cite{risto-vaarandi} for his
\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
be outlined as part of the general explanation below.

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained when writing rules and performs a large number of
        substitutions on the log input lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \IP{} addresses, the
        standard start of rejection messages, etc.) with \regex{} keywords
        that the parser will expand when it loads the rule (see the
        \regex{} entry in \sectionref{rule attributes}).  The purpose of
        this step is to utilise existing knowledge to create more accurate
        \regexes{}.  The new log lines are written to a temporary file,
        which all subsequent stages use instead of the original input file.

@@

        The original algorithm only maintains counters for words whose hash
        result from the previous step has a high frequency; this reduces
        the number of counters maintained by the algorithm, reducing the
        memory requirements of the algorithm.

        As time goes on the amount of memory typically available increases
        and the requirement to reduce memory requirements decreases, hence
        the modified algorithm omits the hashing step and maintains
        counters for all tokens. In addition most uncommon or infrequently
        occurring words will have been substituted with keywords during the
        first step, vastly reducing the number of tuples to maintain
        counters for.

@@

    \item [Build regexes]  The words are reassembled to produce a \regex{}
        matching the log line, and a counter maintained for each \regex{}.
        Contiguous sequences of \texttt{.+} in the newly reassembled
        \regexes{} are collapsed to a single \texttt{.+}; any resulting
        duplicate \regexes{} are combined, and their counters added
        together.  If the \regex{}'s counter is lower than the threshold
        supplied by the user the \regex{} is discarded; this second
        threshold is independent of the threshold used to differentiate
        between fixed and variable words, but once again 1\% of input log
        lines is a good starting point; obviously the threshold depends on
        the number and content of input log lines.  The new \regexes{} are
        displayed for the user to add to the database, either as new rules
        or merged into the \regexes{} of existing rules; the counter for
        each \regex{} is also displayed, giving the user an indication of
        how many of the input log lines that \regex{} should match.
        Discarding \regexes{} will result in some of the input log lines
        not being matched; the utility should be run again once the
        unmatched log lines have been isolated by parsing the original log
        files once the parser has been augmented with the new rules.

@@

A second utility is also provided that reads a list of new \regexes{} and
the input given to the first utility.  It tries to match each input log
line against each \regex{}, counting the number of log lines that match
each \regex{}, warning the user if an input log line is matched by more
than one \regex{}, and additionally warning if an input log line is not
matched by any \regex{}.  It displays a summary of how many input log lines
each \regex{} matched, comparing it to the expected number of matches; this
provides the user with an easy method of checking if the \regexes{}
produced by the first utility are correctly matching the input log lines
they are based upon.  A future version of this utility will also group
input log lines by \regex{}, so the user can tweak the \regexes{} if
required.

@@

Whereas the rules are quite simple and each rule is completely independent
of the other rules, the algorithm is significantly more complicated and
highly internally interdependent.  The algorithm deals with all the
complications of parsing, the eccentricities and oddities of Postfix log
files, presenting the resulting data in a normalised, simple to use
representation.  The algorithm's task is to follow the journey each mail
takes through Postfix, combining the data extracted by rules into a
coherent whole, saving it in a useful and consistent form, and performing
housekeeping duties.

@@

    \item The first set of complications encountered: initially obvious
        difficulties that had to be overcome.

@@

    \item Additional complications that have arisen during the development
        of this parser are documented, the solutions to those
        complications, and where those solutions are implemented in the
        algorithm

@@

\Figureref{flow chart image part 1} and \figureref{flow chart image part 2}
show the paths the data representing a mail can take through the parser
algorithm.  The flow chart covers the most common paths only; there are
additional, uncommon paths excluded for the sake of clarity; details of the
deviations can be found in \sectionref{additional complications}.

@@

    \item The last alternative is that the mail is a parent mail (action:
        COMMIT\@; flowchart:~21).  Regardless of the state of its children
        its data is cleaned up and entered in the database (flowchart:~22).
        The parent may have children that are waiting to be entered in the
        database (flowchart:~23); each of those children's data is cleaned
        up and entered in the database, then deleted from the state tables.
        The parent may also have outstanding children which are not yet
        delivered (flowchart:~24), in which case the parent must wait for
        those children to be finished with.  As soon as the last child is
        deleted from the state tables the parent will also be finished with
        (flowchart:~25), and deleted from the state tables.

@@

The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
that is far more complicated than the behaviour explicitly encoded in the
algorithm.  The top-level parser could be written in pseudo-code like so
(with indentation level indicating flow control):

@@

    \item Four implicit branches, where the transition is determined by the
        next log line that is processed for that mail, e.g.\ state 1 can
        transition to either state 4 or state 5.  (1--\{4,5\},
        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})

@@

            \item [host] The hostname of the server that logged the line.

            \item [program] The name of the program that logged the line.

            \item [pid] The \pid{} of the program that logged the line.

@@

    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
        to deal with unusual situations --- after five hours the timer will
        expire and the \daemon{smtpd} will exit.  This occurs very
        infrequently, as there are many other timeouts that should occur
        in the intervening hours: \DNS{} timeouts, timeouts reading data
        from the client, etc\@.  The active connection for that
        \daemon{smtpd} must be cleaned up, entered in the database, and
        deleted from the state tables.

@@

Adding new actions is not as simple as adding new rules, though care has
been taken in the parser's design and implementation to make adding new
actions as painless as possible.  The implementor writes a subroutine that
accepts the standard arguments given to actions, and registers it as an
action\footnote{The new action must be registered before the rules are
loaded, as it is an error for a rule to reference an unregistered action;
this helps catch mistakes made when adding new rules.} by calling the
framework subroutine add\_actions() with the name of the new action
subroutine as a parameter.  No other work is required on the implementor's
part to integrate the action into the parser; all of their attention and
effort can be focused on the correctness of their action.  The only
negative aspect is that the process involves editing the parser source
code, which makes upgrading to a later version of the parser more
difficult, though by no means impossible.  If the author of the new action
wishes, they can take advantage of the parser's object oriented
implementation~\cite{Wikipedia-object-orientation} by subclassing it and
implementing their changes in the derived class, allowing future upgrading
of the parser with greatly reduced chance of conflicts.\footnote{The real
difference between the two approaches is where the new code is placed.  The
simpler option is to change the parser code directly, but those changes
will then have to be made to subsequent versions of the parser, and as the
scope of the changes increases so does the chance of conflict, or mistakes
when copying the action.  The more time consuming option is to write a
subclass containing the new actions and change the program which invokes
the parser so that it uses the subclass rather than the parser; the changes
required to the program invoking the parser are minor and much less likely
to lead to conflicts when upgrading to future versions of the parser.  An
alternative is to submit new actions to the author of the parser for
inclusion in future versions, resulting in two benefits: the new actions do
not need to be maintained separately, and other users of the parser will
avail of the new functionality.} The action may need to extend the list of
valid combinations described in \sectionref{out of order log lines} if the
addition creates a different set of acceptable programs, but this is
unlikely to occur, as it would require parsing log lines from Postfix
components the parser currently ignores.\footnote{The mail server used for
development does not utilise either the \daemon{lmtp} or \daemon{virtual}
delivery agents, so this parser does not have rules to handle log lines
from those components.  Adding new rules to parse those component's log
lines is a simple process, though if their behaviour differs significantly
from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
required.  The mail server in question is a production mail server handling
mail for a university department; the benefit is that the log files used
exhibit the idiosyncrasies and peculiarities a mail server in the wild must
deal with, but the downside is that significantly altering the
configuration just to log messages from a different Postfix component is
not an option.}

@@

This section presented the core of the parser, starting with a very high
level view and the initial complications that arose.  A flow chart showing
the paths a mail may take through the nascent simplified algorithm was
provided, followed by an explanation of those paths, and a discussion of
the parser's emergent behaviour --- the data from the log files creates the
paths in the flow chart, they are not specified anywhere in the parser.
The framework which holds the parser together was covered next, after which
came a description of the current actions provided by the parser, and the
algorithm for analysing unparsed log lines to create \regexes{} for new
rules.  Detecting, diagnosing and defeating complications forms the largest
single portion of this document, mirroring the development of the parser,
The complications are described in the order they were overcome, with
subsequent problems affecting fewer mails (often by an order or magnitude),
though the time required to solve problems increased with each successive
problem.

@@

Postfix 2.2.x (and presumably previous versions) lacks explicit logging
when bounce notifications are generated; suddenly there will be log entries
for a mail that lacks an obvious source.  There are similarities to the
problem of re-injected mails discussed in \sectionref{tracking re-injected
mail}, but unlike the solution described therein bounce notifications do
not eventually have a log line that identifies their source.  Heuristics
must be used to identify bounce notifications, and those heuristics are:

@@

Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
client aborts the first delivery attempt after the first recipient is
accepted, then makes a second delivery attempt for the same recipient which
it continues with until delivery is complete.\footnote{Microsoft Outlook is
one client that behaves in this fashion; no attempt has been made to
identify other clients that act in a similar way.}  An example dialogue
exhibiting this behaviour is presented below (lines starting with a three
digit number are sent by the server, the other lines are sent by the
client):

@@

    \item If one or more recipients have been accepted Postfix will have
        allocated a queueid for the incoming mail, and there will be a mail
        in the state tables that needs to be dealt with.

@@

The author has only observed this complication occurring after a timeout,
though there may be other circumstances that trigger it.  Sometimes the
\daemon{cleanup} line is logged after the timeout line; parsing this line
causes the creation of a new state table entry for the queueid in the log
line.  This is incorrect because the line actually belongs to the mail
that has just been discarded; the next log line for that queueid will be
seen when the queueid is reused for a different mail, causing a queueid
clash and the appropriate warning.

@@

The solution to this problem is to examine the list of programs that have
logged lines for each mail, comparing the list against a table of
known-good combinations of programs (this check is performed during the
COMMIT action).  If the mail's combination is found in the valid list the
mail can be entered in the database; if the combination is not found entry
must be postponed and the mail flagged for later entry.  The
SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it is
found; if the additional log lines have caused the mail to reach a valid
combination entry will proceed, otherwise it must be postponed once more.

@@

    \item It appears that the hostname used in the HELO command is not
        logged if the mail is accepted.\footnote{Tested with Postfix
        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
        2.5. XXX TEST THIS AGAIN.}  Rectifying this is relatively simple:
        create a restriction that is guaranteed to warn for every accepted
        mail, as follows:
